---
title: "Interpreting Results"
author: "Marcello"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## Introduction

Statistical significance and clinical relevance are not the same thing. A method comparison study may yield statistically significant differences between methods that are clinically negligible, or conversely, reveal clinically important biases that fail to reach statistical significance in small samples.

This vignette focuses on the interpretive aspects of method comparison, helping you translate statistical outputs from `valytics` into meaningful conclusions for your specific application.

```{r load}
library(valytics)
library(ggplot2)
```

## Framework for Interpretation

Before analyzing data, establish your acceptance criteria. These should be based on:

1. **Clinical decision points**: What measurement thresholds trigger clinical action?
2. **Biological variation**: How much does the analyte naturally vary within and between individuals?
3. **Total allowable error**: What combined bias and imprecision is acceptable for the intended use?
4. **Regulatory requirements**: Do applicable guidelines specify performance criteria?

Without predefined criteria, there is a temptation to rationalize whatever results emerge, undermining the objectivity of the comparison.

## Interpreting Bland-Altman Results

### Understanding Bias

The mean difference (bias) quantifies systematic deviation between methods:

```{r ba-example}
data("creatinine_serum")
ba <- ba_analysis(
  x = creatinine_serum$enzymatic,
  y = creatinine_serum$jaffe
)
summary(ba)
```

Key questions to address:

**Is the bias statistically significant?**
The confidence interval for the mean difference tells us whether the observed bias is distinguishable from zero. If the 95% CI excludes zero, the bias is statistically significant at the 0.05 level.

**Is the bias clinically meaningful?**
This depends entirely on context. For creatinine, a bias of 0.1 mg/dL may be trivial at high concentrations but could affect clinical staging of kidney disease near diagnostic thresholds.

Consider expressing bias relative to clinical decision points:

```{r bias-context}
# Example: creatinine threshold for CKD Stage 3 is ~1.2-1.5 mg/dL
bias <- ba$results$bias
cat("Observed bias:", round(bias, 3), "mg/dL\n")
cat("As percentage of CKD threshold (1.3 mg/dL):",
    round(100 * abs(bias) / 1.3, 1), "%\n")
```

### Interpreting Limits of Agreement

The 95% limits of agreement (LoA) define the range within which most differences fall. These are often more clinically relevant than the bias alone.

```{r loa-interpretation}
loa_lower <- ba$results$loa_lower
loa_upper <- ba$results$loa_upper

cat("95% Limits of Agreement:\n")
cat("  Lower:", round(loa_lower, 3), "mg/dL\n")
cat("  Upper:", round(loa_upper, 3), "mg/dL\n")
cat("  Range:", round(loa_upper - loa_lower, 3), "mg/dL\n")
```

Ask yourself: If two methods can differ by as much as the LoA range for the same sample, would this affect clinical decisions?

### Checking Assumptions

The validity of LoA depends on normally distributed differences. Check this with the Shapiro-Wilk test (reported in `summary()`) and visually:

```{r normality-check, fig.cap = "Histogram of differences to assess normality."}
# Extract differences
differences <- ba$results$differences

# Visual check
ggplot(data.frame(diff = differences), aes(x = diff)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15,
                 fill = "steelblue", alpha = 0.7) +
  geom_density(linewidth = 1) +
  labs(x = "Difference (Jaffe - Enzymatic)", y = "Density") +
  theme_minimal()
```

If differences are substantially non-normal, consider:

- Logarithmic transformation for skewed data
- Percentage differences if variability increases with magnitude
- Non-parametric approaches

### Detecting Proportional Bias

A constant bias affects all measurements equally, but proportional bias means the difference depends on the measurement magnitude. The Bland-Altman plot can reveal this:

```{r proportional-bias, fig.cap = "Bland-Altman plot - look for trends indicating proportional bias."}
plot(ba)
```

Signs of proportional bias include:

- Differences that increase (or decrease) systematically with the average value
- A "fan" pattern where variability changes across the measurement range
- Regression of differences on averages with a significant slope

If proportional bias is present, Passing-Bablok regression is particularly informative, as the slope quantifies the proportional relationship.

## Interpreting Passing-Bablok Results

### Slope and Intercept

Passing-Bablok regression addresses two questions directly:

```{r pb-example}
pb <- pb_regression(
  x = creatinine_serum$enzymatic,
  y = creatinine_serum$jaffe
)
summary(pb)
```

**Slope interpretation:**

- Slope = 1: Methods have the same proportional response
- Slope > 1: Method Y gives higher values as concentration increases
- Slope < 1: Method Y gives lower values as concentration increases

The confidence interval is key. If it includes 1, we cannot conclude proportional bias exists.

**Intercept interpretation:**

- Intercept = 0: No constant systematic difference
- Intercept > 0: Method Y consistently reads higher by a fixed amount
- Intercept < 0: Method Y consistently reads lower by a fixed amount

Again, check whether the confidence interval includes zero.

### Clinical Translation

Translate regression parameters into practical terms:

```{r pb-translation}
slope <- pb$results$slope
intercept <- pb$results$intercept

# At a clinical decision point
decision_point <- 1.3  # mg/dL for CKD staging
expected_y <- intercept + slope * decision_point

cat("At enzymatic creatinine =", decision_point, "mg/dL:\n")
cat("Expected Jaffe result:", round(expected_y, 3), "mg/dL\n")
cat("Difference:", round(expected_y - decision_point, 3), "mg/dL\n")
```

This type of calculation shows whether differences at specific clinical thresholds could affect patient classification.

### Linearity Assessment

The CUSUM test evaluates whether the linear model is appropriate:

```{r cusum-check}
cat("CUSUM test p-value:", round(pb$cusum$p_value, 3), "\n")
if (pb$cusum$p_value < 0.05) {
  cat("Significant non-linearity detected.\n")
} else {
  cat("No significant deviation from linearity.\n")
}
```

```{r cusum-plot, fig.cap = "CUSUM plot for linearity assessment."}
plot(pb, type = "cusum")
```

Non-linearity suggests that the relationship between methods changes across the measurement range. This could indicate:

- Matrix effects at certain concentrations
- Interference affecting one method differently across the range
- Calibration issues at the extremes

If non-linearity is present, consider analyzing subranges separately or applying transformations.

## Common Interpretation Pitfalls

### Pitfall 1: Relying on Correlation

Correlation coefficients are frequently reported in method comparison studies but are often misleading:

```{r correlation-pitfall}
# High correlation can coexist with poor agreement
cor_value <- cor(creatinine_serum$enzymatic, creatinine_serum$jaffe)
cat("Correlation coefficient:", round(cor_value, 3), "\n")
```

A high correlation indicates that the methods rank samples similarly, not that they give the same values. Methods with perfect correlation (r = 1) but different calibrations would show systematic bias that correlation fails to detect.

### Pitfall 2: Ignoring Sample Characteristics

Method comparison results depend on the sample population studied. Consider:

- **Concentration range**: Bias may differ at low vs. high concentrations
- **Sample types**: Fresh vs. frozen, arterial vs. venous
- **Patient population**: Healthy volunteers vs. patients with interfering conditions

Results from one population may not generalize to another.

### Pitfall 3: Confusing Statistical and Clinical Significance

```{r significance-example}
# Example with a small but statistically significant bias
n <- 200
set.seed(42)
x <- rnorm(n, mean = 100, sd = 15)
y <- x + rnorm(n, mean = 0.5, sd = 2)  # Small bias of 0.5 units

ba_demo <- ba_analysis(x, y)
cat("Bias:", round(ba_demo$results$bias, 3), "\n")
cat("95% CI for bias:", round(ba_demo$results$bias_ci["lower"], 3), "to",
    round(ba_demo$results$bias_ci["upper"], 3), "\n")

# The bias is statistically significant (CI excludes 0)
# but may be clinically irrelevant depending on the application
```

Always interpret results in the context of what differences matter for patient care or regulatory compliance.

### Pitfall 4: Overfitting Conclusions to Data

Avoid the temptation to explain away unfavorable results or cherry-pick favorable subanalyses. If you established acceptance criteria before the study, stick to them. If results are borderline, acknowledge the uncertainty rather than forcing a definitive conclusion.

## Reporting Recommendations

A complete method comparison report should include:

**Study design:**

- Sample size and power considerations
- Sample selection and handling
- Measurement protocols for both methods

**Statistical results:**

- Bland-Altman: bias with 95% CI, LoA with 95% CI
- Passing-Bablok: slope and intercept with 95% CIs, CUSUM p-value
- Assessment of normality and linearity assumptions

**Visualizations:**

- Bland-Altman plot
- Passing-Bablok scatter with regression line
- CUSUM plot if linearity is questioned

**Interpretation:**

- Comparison to predefined acceptance criteria
- Clinical implications at relevant decision points
- Limitations and generalizability

```{r report-example}
# Example summary for a report
cat("=== Method Comparison Summary ===\n\n")
cat("Bland-Altman Analysis:\n")
cat("  Bias:", round(ba$results$bias, 3), "mg/dL",
    "(95% CI:", round(ba$results$bias_ci["lower"], 3), "to",
    round(ba$results$bias_ci["upper"], 3), ")\n")
cat("  LoA:", round(ba$results$loa_lower, 3), "to",
    round(ba$results$loa_upper, 3), "mg/dL\n\n")

cat("Passing-Bablok Regression:\n")
cat("  Slope:", round(pb$results$slope, 3),
    "(95% CI:", round(pb$results$slope_ci["lower"], 3), "to",
    round(pb$results$slope_ci["upper"], 3), ")\n")
cat("  Intercept:", round(pb$results$intercept, 3),
    "(95% CI:", round(pb$results$intercept_ci["lower"], 3), "to",
    round(pb$results$intercept_ci["upper"], 3), ")\n")
cat("  CUSUM p-value:", round(pb$cusum$p_value, 3), "\n")
```

## Worked Example: Troponin Comparison

Let us apply these principles to the cardiac troponin dataset, which presents additional challenges due to its wide dynamic range:

```{r troponin-analysis}
data("troponin_cardiac")

# Bland-Altman with percentage differences (appropriate for wide range)
ba_trop <- ba_analysis(
  x = troponin_cardiac$platform_a,
  y = troponin_cardiac$platform_b,
  type = "percent"
)
summary(ba_trop)
```

```{r troponin-ba-plot, fig.cap = "Bland-Altman plot for troponin (percentage differences)."}
plot(ba_trop)
```

```{r troponin-pb}
pb_trop <- pb_regression(
  x = troponin_cardiac$platform_a,
  y = troponin_cardiac$platform_b
)
summary(pb_trop)
```

```{r troponin-pb-plot, fig.cap = "Passing-Bablok regression for troponin platforms."}
plot(pb_trop, type = "scatter")
```

For troponin, clinical interpretation must consider:

- The 99th percentile upper reference limit (diagnostic threshold)
- Differences near this threshold that could affect acute MI diagnosis
- Whether proportional differences are acceptable given the wide measurement range

## Conclusion

Rigorous interpretation of method comparison results requires:

1. Predefined, clinically-grounded acceptance criteria
2. Understanding of both statistical and clinical significance
3. Careful assessment of assumptions underlying the statistical methods
4. Transparent reporting of limitations

The tools in `valytics` provide the statistical foundation; your domain expertise transforms those numbers into actionable conclusions.

## References

Bland JM, Altman DG. Statistical methods for assessing agreement between two methods of clinical measurement. Lancet. 1986;1(8476):307-310.

Westgard JO, Hunt MR. Use and interpretation of common statistical tests in method-comparison studies. Clinical Chemistry. 1973;19(1):49-57.

Linnet K. Necessary sample size for method comparison studies based on regression analysis. Clinical Chemistry. 1999;45(6):882-894.

Passing H, Bablok W. A new biometrical procedure for testing the equality of measurements from two different analytical methods. Journal of Clinical Chemistry and Clinical Biochemistry. 1983;21(11):709-720.

CLSI. Measurement Procedure Comparison and Bias Estimation Using Patient Samples; Approved Guideline. EP09-A3. Clinical and Laboratory Standards Institute; 2013.
