---
title: "Understanding Method Comparison Statistics"
author: "Marcello Grassi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding Method Comparison Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## Introduction

This vignette provides a conceptual overview of the statistical methods implemented in `valytics`. The goal is to help you understand *what* the numbers mean and *how* to think about them, not to prescribe specific acceptance criteria or make decisions for you.

Whether your analysis "passes" or "fails" depends entirely on your specific application, regulatory requirements, and clinical context. This package provides the tools; you and your organization define what constitutes acceptable agreement.

```{r load}
library(valytics)
library(ggplot2)
```

## Statistical Concepts in Bland-Altman Analysis

### What is Bias?

The **bias** (mean difference) quantifies the average systematic offset between two methods. It answers: "On average, how much higher or lower does method Y read compared to method X?"

```{r ba-example}
data("creatinine_serum")
ba <- ba_analysis(
  x = creatinine_serum$enzymatic,
  y = creatinine_serum$jaffe
)
```

```{r bias-output}
cat("Bias:", round(ba$results$bias, 3), "mg/dL\n")
cat("95% CI:", round(ba$results$bias_ci["lower"], 3), "to",
    round(ba$results$bias_ci["upper"], 3), "\n")
```

**What this tells you:**

- The point estimate indicates the direction and magnitude of systematic difference
- The confidence interval quantifies uncertainty due to sampling variability
- If the CI excludes zero, the bias is statistically distinguishable from zero at that confidence level

**What this does NOT tell you:**

- Whether the bias is clinically important (that depends on your application)
- Whether the methods are "equivalent" (you must define what that means)
- Whether you should use one method over another

### What are Limits of Agreement?

The **limits of agreement (LoA)** define an interval expected to contain 95% of the differences between methods. They answer: "For a randomly selected sample, how much could the two methods disagree?"

```{r loa-output}
cat("Lower LoA:", round(ba$results$loa_lower, 3), "\n")
cat("Upper LoA:", round(ba$results$loa_upper, 3), "\n")
cat("Width:", round(ba$results$loa_upper - ba$results$loa_lower, 3), "\n")
```

The LoA represent the *range* of disagreement you can expect in practice. A narrow LoA indicates consistent agreement; a wide LoA indicates variable differences.

**Key insight:** The LoA are often more informative than the bias alone. Two methods might have negligible average bias but wide limits of agreement, meaning individual measurements could differ substantially.

### Visualizing Agreement

The Bland-Altman plot provides a visual assessment:

```{r ba-plot, fig.cap = "Bland-Altman plot showing differences vs. averages."}
plot(ba)
```

**What to look for:**

- **Random scatter around the bias line**: Suggests constant bias across the measurement range
- **Funnel shape**: Variance changes with magnitude (heteroscedasticity)
- **Systematic trend**: Proportional bias (differences depend on concentration)
- **Points outside LoA**: Expected for ~5% of observations if assumptions hold

### Checking Assumptions

Bland-Altman analysis assumes normally distributed differences. The summary provides a Shapiro-Wilk test:

```{r normality}
summ <- summary(ba)
if (!is.null(summ$normality_test)) {
  cat("Shapiro-Wilk p-value:", round(summ$normality_test$p.value, 4), "\n")
}
```

A low p-value suggests non-normality. Consider:

- Examining the distribution visually
- Using percentage differences if variance increases with magnitude
- Applying transformations for skewed data

```{r histogram, fig.cap = "Distribution of differences."}
ggplot(data.frame(diff = ba$results$differences), aes(x = diff)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15,
                 fill = "steelblue", alpha = 0.7) +
  geom_density(linewidth = 1) +
  labs(x = "Difference (Jaffe - Enzymatic)", y = "Density") +
  theme_minimal()
```

## Statistical Concepts in Passing-Bablok Regression

### Slope and Intercept

Passing-Bablok regression fits a line: `Y = intercept + slope * X`

The parameters have direct interpretations:

- **Slope = 1**: No proportional (multiplicative) difference
- **Intercept = 0**: No constant (additive) difference

```{r pb-example}
pb <- pb_regression(
  x = creatinine_serum$enzymatic,
  y = creatinine_serum$jaffe
)
```

```{r pb-output}
cat("Slope:", round(pb$results$slope, 4), "\n")
cat("  95% CI:", round(pb$results$slope_ci["lower"], 4), "to",
    round(pb$results$slope_ci["upper"], 4), "\n")
cat("Intercept:", round(pb$results$intercept, 4), "\n")
cat("  95% CI:", round(pb$results$intercept_ci["lower"], 4), "to",
    round(pb$results$intercept_ci["upper"], 4), "\n")
```

**How to read the confidence intervals:**

- If the slope CI **includes 1**: Cannot conclude proportional bias exists
- If the slope CI **excludes 1**: Evidence of proportional bias
- If the intercept CI **includes 0**: Cannot conclude constant bias exists
- If the intercept CI **excludes 0**: Evidence of constant bias

### Translating to Practical Differences

You can use the regression equation to estimate expected differences at specific concentrations:

```{r translation}
# At various concentrations, what's the expected difference?
concentrations <- c(0.8, 1.3, 3.0, 6.0)

for (conc in concentrations) {
  expected_y <- pb$results$intercept + pb$results$slope * conc
  difference <- expected_y - conc
  cat(sprintf("At X = %.1f: expected Y = %.3f, difference = %.3f\n",
              conc, expected_y, difference))
}
```

This helps translate abstract regression parameters into concrete, application-specific terms.

### Linearity Assessment

The CUSUM test evaluates whether a linear model is appropriate:

```{r cusum}
cat("CUSUM statistic:", round(pb$cusum$statistic, 4), "\n")
cat("p-value:", round(pb$cusum$p_value, 4), "\n")
```

A significant result (conventionally p < 0.05) suggests the relationship may not be linear across the measurement range. If non-linearity is detected:

- Consider whether it's clinically meaningful
- Examine specific concentration ranges
- Evaluate whether a single regression is appropriate

```{r cusum-plot, fig.cap = "CUSUM plot for linearity assessment."}
plot(pb, type = "cusum")
```

## Common Analysis Considerations

### Correlation is Not Agreement

High correlation between methods is often reported but can be misleading:

```{r correlation}
r <- cor(creatinine_serum$enzymatic, creatinine_serum$jaffe)
cat("Correlation coefficient:", round(r, 4), "\n")
```

Correlation measures whether methods *rank* samples similarly, not whether they give the *same values*. Two methods with r = 1 but different calibrations would show systematic bias that correlation fails to detect.

### Sample Characteristics Matter

Your results depend on:

- **Concentration range**: Bias may differ at low vs. high concentrations
- **Sample types**: Matrix effects can vary
- **Population**: Results from one patient group may not generalize

Be cautious about extrapolating beyond the conditions of your study.

### Statistical vs. Practical Significance

A statistically significant bias (CI excludes zero) may or may not be practically important. Consider:

```{r context}
# Example: Is a bias of X clinically meaningful?
# This depends entirely on YOUR application
bias_value <- ba$results$bias

cat("Observed bias:", round(bias_value, 3), "mg/dL\n")
cat("\nWhether this is 'acceptable' depends on:\n")
cat("- Your specific clinical decision thresholds\n")
cat("- Regulatory requirements for your application\n")
cat("- Intended use of the measurement\n")
cat("- Established performance goals (CLIA, biological variation, etc.)\n")
```

## Creating Analysis Reports

Here's how to extract key statistics for reporting:

```{r report}
# Bland-Altman summary
cat("=== Bland-Altman Analysis ===\n")
cat(sprintf("n = %d\n", ba$input$n))
cat(sprintf("Bias: %.3f (95%% CI: %.3f to %.3f)\n",
            ba$results$bias,
            ba$results$bias_ci["lower"],
            ba$results$bias_ci["upper"]))
cat(sprintf("SD of differences: %.3f\n", ba$results$sd_diff))
cat(sprintf("LoA: %.3f to %.3f\n\n",
            ba$results$loa_lower,
            ba$results$loa_upper))

# Passing-Bablok summary
cat("=== Passing-Bablok Regression ===\n")
cat(sprintf("Slope: %.4f (95%% CI: %.4f to %.4f)\n",
            pb$results$slope,
            pb$results$slope_ci["lower"],
            pb$results$slope_ci["upper"]))
cat(sprintf("Intercept: %.4f (95%% CI: %.4f to %.4f)\n",
            pb$results$intercept,
            pb$results$intercept_ci["lower"],
            pb$results$intercept_ci["upper"]))
cat(sprintf("CUSUM p-value: %.4f\n", pb$cusum$p_value))
```

## Summary

The `valytics` package provides statistical tools for method comparison. It calculates:

- **Bland-Altman**: Bias, limits of agreement, and their confidence intervals
- **Passing-Bablok**: Slope, intercept, and linearity assessment

These statistics describe the *relationship* between methods. Whether that relationship is "acceptable" for your purpose is a separate question that depends on:

- Clinical decision thresholds
- Regulatory requirements
- Performance specifications (biological variation, CLIA, etc.)
- Intended use

The package reports what the data show. You decide what it means for your application.

## References

Bland JM, Altman DG. Statistical methods for assessing agreement between two methods of clinical measurement. Lancet. 1986;1(8476):307-310.

Bland JM, Altman DG. Measuring agreement in method comparison studies. Statistical Methods in Medical Research. 1999;8(2):135-160.

Passing H, Bablok W. A new biometrical procedure for testing the equality of measurements from two different analytical methods. Journal of Clinical Chemistry and Clinical Biochemistry. 1983;21(11):709-720.

Westgard JO, Hunt MR. Use and interpretation of common statistical tests in method-comparison studies. Clinical Chemistry. 1973;19(1):49-57.
